### 2022-08-25 article.Rmd
First few thoughts

---
title: "Two years of CiTO annotations"
author:
- Egon Willighagen [1,*,orcid:0000-0001-7542-0286]
link-citations: yes
output:
  html_document:
    pandoc_args:
    - --lua-filter=filters/extract-cito.lua
    - --citeproc
    - --lua-filter=filters/insert-cito-in-ref.lua
  pdf_document:
    pandoc_args:
    - --lua-filter=filters/extract-cito.lua
    - --citeproc
    - --lua-filter=filters/insert-cito-in-ref.lua
  word_document:
    pandoc_args:
    - --lua-filter=filters/extract-cito.lua
    - --citeproc
    - --lua-filter=filters/insert-cito-in-ref.lua
# RStudio: cito-bibliography: bibliography.bib
# Other: bibliography: bibliography.bib
cito-bibliography: bibliography.bib
csl: journal-of-cheminformatics.csl
...

* Correspondance:
egon.willighagen@maastrichtuniversity.nl

# Abstract

Citations are an essential aspect of research communication and have become the basis of many
metrics in the academic publishing world. Some see citation counts as a mark of scientific impact or
even quality, but in reality the reasons for citing other work are manifold.
Two years ago the *Journal of Cheminformatics* proposed a pilot for the adoption of annotating
citations with their annotations. Basically, when you cite a source like a journal
article or datasets, you also explain why specifically you cite that article. Particularly,
the agreement and disagreement and reuse of methods and data is interesting.
This article explores what happened after the launch of the pilot, gives examples of
how people used the idea, and what needs to happen next.

# Main text

Communiting new research findings is primarily done by written texts in the form of scholarly
articles, books, and book chapters. To not having to repeat past research by themselves or others,
authors cite research relevant [@agreesWith:citesAsAuthority:Shotton_2010]. However, the reasons why
authors cite literature vary, which complicates how we use citations [@agreesWith:citesAsAuthority:Nicholson_2021]. 
Typing citations is therefore of interest: it allows us to navigate ltierature more easily: it points
us to essential research methods, data, and can warn us of research that cannot be reproduced,
or others disagree with [@Nicholson_2021].

With the use of citations increasingly being picked up to help researchers with tools like
scite.ai [@citesAsAuthority:Nicholson_2021] en Connected Papers, having typed
citations will help us explore literature. Therefore, the *Journal of Cheminformatics*
started a pilot for the adoption of annotating citations with their annotations [@discusses:Willighagen_2020].

# The Citation Typing Ontology Pilot

The pilot consists of a couple of components and the editorial explains some of them [@Willighagen_2020].
The Citation Typing Ontology was selected to express the intention [@usesMethodIn:Shotton_2010],
the intention is expressed a compact identifier wrapped in square brackets [@usesMethodIn:Wimalaratne_2018], and 
the *bibnotes* concept of the Springer Nature publishing platform was used as carrier.
Authors are guided by a landing page consisting of a BioMedCentral Collection at
[https://www.biomedcentral.com/collections/cito](https://www.biomedcentral.com/collections/cito)
and author guidelines explaining to authors
how they can add the annotations with their favoriate editor at
[https://jcheminform.github.io/jcheminform-author-guidelines/cito](https://jcheminform.github.io/jcheminform-author-guidelines/cito).

Because the CiTO ontology has many terms for many different citation intentions,
we made a selection of CiTO terms authors could use [@CiTOCollection]:
**[cito:citesAsDatasource]** to indicate
to a source the provides data to back up a claim, **[cito:usesDataFrom]** to indicate that the authors
reused data, **[cito:usesMethodIn]** when a method or protocol explain in that source is used,
and a few more general intentions like **[cito:discusses]**, **[cito:extends]**, **[cito:agreesWith]**,
and **[cito:disagreesWith]**.
The journal itself would adopt the following CiTO annotations: **[cito:retracts]**,
**[cito:repliesTo]**, and **[updates]**. Fortunately, it has not been used yet, but
the first would be used if an article with be retracted from the journal. The second
would be used when an Letter to the Editor replies to an earlier published article,
and **[cito:updates]** when an Correction was published.

## Wikidata and Scholia

To track the uptake but also to demonstrate the impact, we extended Scholia to visualize
citation intention data. Scholia is a graphical interface around the data stored in
Wikidata [@usesMethodIn:Nielsen_2017] and includes citations from OpenCitations [@Peroni_2020]
and PubMed. Wikidata allows adding qualifier to statements which allowed us to define
a data model for citations annoted with CiTO intention; the Wikidata property P3712
has been used for this, labeled *objective of project or action* (see Figure 1).

![](images/datamodel.png)
**Figure 1**: Screenshot of the citation statements for an article where the
*objective of project or action* qualifier is used to annotate the citation with their
CiTO intentions.

This data in Wikidata can then be accessed via Scholia and Scholia tells use some
overall statisticss of the number of annotations, which we reported on about a year
ago too [@citesAsDataSource:Willighagen_2021]. Since last year and written down
on August 25, the number of annotations and the number of annotated citations
have almost doubled (from 377 to 603 and from 304 to 494, respectively). The first
number is higher because one citation can have more than one citation intention.
To continue, the current number of citations are citing 387 articles in 141 different
scholarly journals, and they are found in 98 articles in 48 different journals
(see Figure 2) [@citesAsDataSource:scholiaCiTOStatistics].

![](images/scholiaStats.png)
**Figure 2**: XXXX.

It must be noted that the *Journal of Cheminformatics* is only one possible source,
but is still the only journal that uses CiTO annotation explicitly in the articles itself.
And with 335 annotated citations in 32 articles it also is the major source of
CiTO annotations in Wikidata at the time of writing.
However, CiTO intention annotations in Wikidata can come from other sources too
and be added both manually and automatically using the tools around Wikidata.
When all annotation is combined, Scholia shows us that **[cito:citesAsAuthority]**
is the most used intention, with 226 annotated citations (out of 603) in 38 articles.
**[cito:usesMethodIn]** follows with 102 annotated citations.

## Adoption by *Journal of Cheminformatics* authors

XXXXX

## Technological innovation

Markdown template...

To make life easier for authors, and following a Twitter discussion in Spring 2021, we developed a Markdown template with native CiTO support. Here, the author merely indicates the CiTO type when they cite the article.

The manuscript can then easily be saved as a Word file to enable submission to the journal. The *Journal of Cheminformatics* template is available from our GitHub repository, and authors and editors should feel free to adapt it to their own journal needs.

## What is next?

With X articles published in the CiTO Collection, the adoption is not as high as one would have hoped.
Nevertheless, the pilot is triggering interest with both authors and publishing platforms. The Scholia
use case shows how the information can be used downstream. Further uptake of this idea of typed citations
depends on the combined willingness of journal editors, authors, publishers and indexing services alike.
The rise of services like scite.ai shows that the research community needs this kind of information.

#### Availability of data and materials
Text for this section.

#### Competing interests
Text for this section.

#### Funding
Text for this section.

#### Authors' contributions
Text for this section.

#### Acknowledgements

This work would not be possible without the support from Springer Nature and
Matthew Smyllie in particular and the editors of the *Journal of Cheminformatics*,
Rajarshi Guha, Nina Jealizakova, and Barbara Zdrazil. XXX is thanked for
integration the sections on CiTO use into the main Scholia journal aspects [REF].
Huge thanks goes to YYY for developing the Markdown/Pandoc integration.

# References

### 2022-08-26 article.Rmd
Started reporting/summarizing the Results

@@
 link-citations: yes
+geometry: margin=2.5cm
 output:
@@
 
-XXXXX
+|**Year** |**Type**  |**Article** |**Intentions** | **% Annotated citaions** |
+--- |----- |------------ |------------ |---
+|Year 1|Editorial|Adoption of the Citation Typing Ontology by the *Journal of Cheminformatics* [@citesForInformation:Willighagen_2020]|cito:citesAsAuthority, cito:citesForInformation, cito:usesMethodIn||
+||Methodology|Predicting target profiles with confidence as a service using docking scores [@citesForInformation:Ahmed2020]|cito:agreesWith, cito:extends, cito:citesAsAuthority, cito:citesAsDataSource, cito:usesMethodIn|
+||Research article|Too sweet: cheminformatics for deglycosylation in natural products [@citesForInformation:Schaub2020]|cito:citeAsAuthority, cito:citesForInformation, cito:usesDataFrom, cito:usesMethodIn||
+||Educational|A ligand-based computational drug repurposing pipeline using KNIME and Programmatic Data Access: case studies for rare diseases and COVID-19 [@citesForInformation:Tuerkova2020]|cito:agreesWith, cito:citesAsAuthority, cito:citesAsDataSource, cito:discusses, cito:usesDataFrom, cito:usesMethodIn||
 

### 2022-08-26 article.Rmd
Year 1 data collection finished

@@
 
-|**Year** |**Type**  |**Article** |**Intentions** | **% Annotated citaions** |
---- |----- |------------ |------------ |---
-|Year 1|Editorial|Adoption of the Citation Typing Ontology by the *Journal of Cheminformatics* [@citesForInformation:Willighagen_2020]|cito:citesAsAuthority, cito:citesForInformation, cito:usesMethodIn||
-||Methodology|Predicting target profiles with confidence as a service using docking scores [@citesForInformation:Ahmed2020]|cito:agreesWith, cito:extends, cito:citesAsAuthority, cito:citesAsDataSource, cito:usesMethodIn|
-||Research article|Too sweet: cheminformatics for deglycosylation in natural products [@citesForInformation:Schaub2020]|cito:citeAsAuthority, cito:citesForInformation, cito:usesDataFrom, cito:usesMethodIn||
-||Educational|A ligand-based computational drug repurposing pipeline using KNIME and Programmatic Data Access: case studies for rare diseases and COVID-19 [@citesForInformation:Tuerkova2020]|cito:agreesWith, cito:citesAsAuthority, cito:citesAsDataSource, cito:discusses, cito:usesDataFrom, cito:usesMethodIn||
+**Table 1**: *Journal of Cheminformatics* articles in the first year of the pilot.
+
+|**Type**  |**Article** |**Intentions** | **%CiTO** |
+----- |------------ |------------ |---
+|Editorial|Adoption of the Citation Typing Ontology by the *Journal of Cheminformatics* [@citesForInformation:Willighagen_2020]|cito:citesAsAuthority, cito:citesForInformation, cito:usesMethodIn|42%|
+|Methodology|Predicting target profiles with confidence as a service using docking scores [@citesForInformation:Ahmed2020]|cito:agreesWith, cito:extends, cito:citesAsAuthority, cito:citesAsDataSource, cito:usesMethodIn|47%|
+|Research article|Too sweet: cheminformatics for deglycosylation in natural products [@citesForInformation:Schaub2020]|cito:citeAsAuthority, cito:citesForInformation, cito:usesDataFrom, cito:usesMethodIn||
+|Educational|A ligand-based computational drug repurposing pipeline using KNIME and Programmatic Data Access: case studies for rare diseases and COVID-19 [@citesForInformation:Tuerkova2020]|cito:agreesWith, cito:citesAsAuthority, cito:citesAsDataSource, cito:discusses, cito:usesDataFrom, cito:usesMethodIn|98%|
+|Database|COCONUT online: Collection of Open Natural Products database [@citesForInformation:Sorokina2021]|cito:citesForInformation, cito:usesDataFrom, cito:usesMethodIn|41%|
+|Editorial|What is the role of cheminformatics in a pandemic?
+[@citesForInformation:Guha2021]|cito:agreesWith, cito:citesAsAuthority, cito:citesAsDataSource, cito:citesForInformation, cito:citesAsPotentialSolution|91%|
+|Research article|Empowering large chemical knowledge bases for exposomics: PubChemLite meets MetFrag [@citesForInformation:Schymanski2021]|cito:citesAsAuthority, cito:citesAsDataSource, cito:citesAsMetadataDocument, cito:discusses, cito:extends, cito:usesDataFrom, cito:usesMethodIn|100%|
+|Software|IDSM ChemWebRDF: SPARQLing small-molecule datasets [@citesForInformation:Galgonek2021]|cito:citesAsAuthority, cito:citesAsRelated, cito:usesDataFrom, cito:usesMethodIn|97%|
+|Letter to the Editor|FAIR chemical structures in the Journal of Cheminformatics [@citesForInformation:SchymanskiBolton2021]|cito:citesAsAuthority, cito:documents|100%|
+|Letter Response|Reply to “FAIR chemical structure in the Journal of Cheminformatics”|cito:agreesWith, cito:citesAsAuthority, cito:citesAsDataSource, cito:obtainsBackgroundFrom, cito:repliesTo| 100%|
 
@@
 The *Journal of Cheminformatics* template is available from our GitHub repository, and authors and editors should feel free to adapt it to their own journal needs.
+
+## Discussion
+
+Formatting issues
+
+Citation level intention annotation: only Markdown
 

### 2022-08-26 article.Rmd
Year 2 done too now

@@
 |Letter to the Editor|FAIR chemical structures in the Journal of Cheminformatics [@citesForInformation:SchymanskiBolton2021]|cito:citesAsAuthority, cito:documents|100%|
-|Letter Response|Reply to “FAIR chemical structure in the Journal of Cheminformatics”|cito:agreesWith, cito:citesAsAuthority, cito:citesAsDataSource, cito:obtainsBackgroundFrom, cito:repliesTo| 100%|
+|Letter Response|Reply to “FAIR chemical structure in the Journal of Cheminformatics” [@citesForInformation:Guha2021b]|cito:agreesWith, cito:citesAsAuthority, cito:citesAsDataSource, cito:obtainsBackgroundFrom, cito:repliesTo| 100%|
+
+**Table 2**: *Journal of Cheminformatics* articles in the second year of the pilot. The * indicates that
+the percentage is based on CiTO intends different from cito:cites.
+
+|**Type**  |**Article** |**Intentions** | **%CiTO** |
+----- |------------ |------------ |---
+|Research article|DECIMER 1.0: deep learning for chemical image recognition using transformers [@citesForInformation:Rajan2021]|||
+|Research article|*DrugEx* v2: de novo design of drug molecules by Pareto-based multi-objective reinforcement learning in polypharmacology[@citesForInformation:Liu2021]|||
+|Database|PSnpBind: a database of mutated binding site protein–ligand complexes constructed using a multithreaded virtual screening workflow [@citesForInformation:Ammar2022]|||
+|Editorial|Diversifying cheminformatics [@citesForInformation:Zdrazil2022]|||
+|Data Note|DECIMER - hand-drawn molecule images dataset [@citesForInformation:Brinkhaus2022]|cito:agreesWith, cito:cites, cito:citesAsAuthority, cito:extends, cito:usesDataFrom, cito:usesMethodIn|41% (*)|
+
+[//]: <> (|Research article|European Registry of Materials: global, unique identifiers for (undisclosed) nanomaterials [@citesForInformation:vanRijn2022]||| --> THIRD YEAR!)
 
@@
 Here, the author merely indicates the CiTO type when they cite the article.
+This is using a method introduced by Krewinkel et al.
+[@usesMethodIn:Krewinkel2017].
 
@@
 Matthew Smyllie in particular and the editors of the *Journal of Cheminformatics*,
-Rajarshi Guha, Nina Jealizakova, and Barbara Zdrazil. XXX is thanked for
-integration the sections on CiTO use into the main Scholia journal aspects [REF].
-Huge thanks goes to YYY for developing the Markdown/Pandoc integration.
+Rajarshi Guha, Nina Jealizakova, and Barbara Zdrazil. Carlin MacKenzie is thanked for
+integration the sections on CiTO use into the main Scholia journal aspects [@citesAsEvidence:0af60cf7732bc8664d828cbe51f233aa63201df9].
+Huge thanks goes to Albert Krewinkel for developing the Markdown/Pandoc integration.
+Finally, thanks to all authors for including CiTO annotation in their articles.
 

### 2022-08-28 article.Rmd
Added the Connected Papers reference, see https://twitter.com/ConnectedPapers/status/1563477087541170176

@@
 With the use of citations increasingly being picked up to help researchers with tools like
-scite.ai [@citesAsAuthority:Nicholson_2021] en Connected Papers, having typed
+scite.ai [@citesAsAuthority:Nicholson_2021] en Connected Papers [@citesAsAuthority:ConnectedPapers], having typed
 citations will help us explore literature. Therefore, the *Journal of Cheminformatics*

### 2022-08-28 article.Rmd
Added data about CiTO use of 2nd year articles

@@
 ----- |------------ |------------ |---
-|Research article|DECIMER 1.0: deep learning for chemical image recognition using transformers [@citesForInformation:Rajan2021]|||
-|Research article|*DrugEx* v2: de novo design of drug molecules by Pareto-based multi-objective reinforcement learning in polypharmacology[@citesForInformation:Liu2021]|||
-|Database|PSnpBind: a database of mutated binding site protein–ligand complexes constructed using a multithreaded virtual screening workflow [@citesForInformation:Ammar2022]|||
-|Editorial|Diversifying cheminformatics [@citesForInformation:Zdrazil2022]|||
+|Research article|DECIMER 1.0: deep learning for chemical image recognition using transformers [@citesForInformation:Rajan2021]|cito:agreesWith, cito:citesAsAuthority, cito:citesAsDataSource, cito:extends, cito:usesMethodIn|66% (*)|
+|Research article|*DrugEx* v2: de novo design of drug molecules by Pareto-based multi-objective reinforcement learning in polypharmacology[@citesForInformation:Liu2021]|cito:extends, cito:citesAsAuthority, cito:usesMethodIn|100%|
+|Database|PSnpBind: a database of mutated binding site protein–ligand complexes constructed using a multithreaded virtual screening workflow [@citesForInformation:Ammar2022]|cito:citesAsDataSource, cito:usesDataFrom, cito:usesMethodIn|23%|
+|Editorial|Diversifying cheminformatics [@citesForInformation:Zdrazil2022]|cito:citesAsAuthority, cito:citesAsEvidence, cito:citesForInformation, cito:citesAsRecommendedReading, cito:citesAsSourceDocument, cito:containsAssertionFrom|100%|
 |Data Note|DECIMER - hand-drawn molecule images dataset [@citesForInformation:Brinkhaus2022]|cito:agreesWith, cito:cites, cito:citesAsAuthority, cito:extends, cito:usesDataFrom, cito:usesMethodIn|41% (*)|

### 2022-08-28 article.Rmd
Wrote up the content of Table 1 and 2

@@
 title: "Two years of CiTO annotations"
+subtitle: "Type: Correspondance"
 author:
@@
 
+In the first two years after the start of the Pilot, including the seminal editorial [@extends:Willighagen_2020],
+the *Journal of Cheminformatics* has published fifteen articles with explicit CiTO annotation:
+three Editorials, four Research articles, two Database, and one of each of Data Note, Software,
+Letter to the Editor, Letter Response, Educational, and Methodology. Ten were published in
+the first year (Table 1) and five in the second year (Table 2). Each article annotated
+one or more citations with CiTO intentions, and several articles annotated every citation,
+far exceeding the original anticipation.
+
+Also exceeding expectations is the diversity of the chosen CiTO intention types. The original
+guidance focused on **[cito:citesAsDataSource]**, **[cito:usesDataFrom]** (the first is used
+to cite an article with data, the second when you reused data and citate the article where it
+comes from), **[cito:usesMethodIn]**, **[cito:citesAsAuthority]**, **[cito:discusses]**,
+**[cito:extends]**, **[cito:agreesWith]**, and **[cito:disagreesWith]**. Not only have
+all of these been used, authors also used **[cito:citesForInformation]**,
+**[cito:citesAsPotentialSolution]**, **[cito:citesAsRelated]**, **[cito:documents]**,
+and **[cito:obtainsBackgroundFrom]**.
+
 **Table 1**: *Journal of Cheminformatics* articles in the first year of the pilot.

### 2022-08-28 article.Rmd
Proper markdown for figure captions

@@
 
-![](images/datamodel.png)
-**Figure 1**: Screenshot of the citation statements for an article where the
+![Screenshot of the citation statements for an article where the
 *objective of project or action* qualifier is used to annotate the citation with their
-CiTO intentions.
+CiTO intentions.](images/datamodel.png)
 
@@
 
-![](images/scholiaStats.png)
-**Figure 2**: XXXX.
+![Screenshot of the Scholia Citation Typing Ontology page showing the
+daily statistics.](images/scholiaStats.png)
 

### 2022-08-28 article.Rmd
Table captions done right too now

@@
 
-**Table 1**: *Journal of Cheminformatics* articles in the first year of the pilot.
-
 |**Type**  |**Article** |**Intentions** | **%CiTO** |
@@
 
-**Table 2**: *Journal of Cheminformatics* articles in the second year of the pilot. The * indicates that
-the percentage is based on CiTO intends different from cito:cites.
+Table: *Journal of Cheminformatics* articles in the first year of the pilot.
 
@@
 |Data Note|DECIMER - hand-drawn molecule images dataset [@citesForInformation:Brinkhaus2022]|cito:agreesWith, cito:cites, cito:citesAsAuthority, cito:extends, cito:usesDataFrom, cito:usesMethodIn|41% (*)|
+
+Table: *Journal of Cheminformatics* articles in the second year of the pilot. The * indicates that
+the percentage is based on CiTO intends different from cito:cites.
 

### 2022-11-13 article.Rmd
Filled in some gaps.

@@
 
-[//]: <> (|Research article|European Registry of Materials: global, unique identifiers for (undisclosed) nanomaterials [@citesForInformation:vanRijn2022]||| --> THIRD YEAR!)
-
 ## Technological innovation
 
-Markdown template...
-
-To make life easier for authors, and following a Twitter discussion in Spring 2021, we developed a Markdown template with native CiTO support.
+To make life easier for authors, and following a Twitter discussion in Spring 2021, a Markdown template was developed with native CiTO support: https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html.
 Here, the author merely indicates the CiTO type when they cite the article.
@@
 [@usesMethodIn:Krewinkel2017].
+The manuscript can then be converted to a Microsoft Word file with Pandoc (https://pandoc.org/) for submission to the journal.
 
-The manuscript can then easily be saved as a Word file to enable submission to the journal.
-The *Journal of Cheminformatics* template is available from our GitHub repository, and authors and editors should feel free to adapt it to their own journal needs.
+The *Journal of Cheminformatics* template is available from the journal's GitHub organization, and authors and editors should feel free to adapt it to the needs of other journals.
+The BioHackrXiv (https://biohackrxiv.org/) preprint server also support CiTO annotations [@cito:citesAsDataSource:Prins2021] and this template can be found at https://github.com/biohackrxiv/publication-template.
 
@@
 
-Formatting issues
+Because many publishing platforms currently do not support display of citation level intention annotation,
+the default model only provides the annotation in the bibliography. The choice was made to use the `bibnotes`
+idea which allows giving additional notes to a reference in the bibliography. The CiTO Pilot uses
+the *safe CURIE* standard proposed by the W3C in 2010 [@usesMethodIn:BirBeck2010]. The Markdown templates
+implement this idea. However, typesetting in the *Journal of Cheminformatics* has shown to be inconsisten,
+even within single articles, suggesting additional training or guidance could be needed.
 
-Citation level intention annotation: only Markdown
+This links to a limitation to the current use of CiTO annotation: the citation-level annotation may be
+supported by some authoring systems (Markdown), even then the depiction may not be supported. If we convert
+a Markdown file to a Microsoft Word file, the annotation is kept in the bibliography. However, when
+writing a manuscript in Microsoft Word, LibreOffice, or Google Docs, combined with a reference manager
+like Zotero, the CiTO annotation cannot be stored as part of the manuscript easily. The problem here is
+that CiTO annotation in the reference manager are no longer linked to when they are cited. The workaround
+is to add the CiTO annotation after completion of the Word document, directly to the bibliography.
+
+LaTeX users may find them in a situation between that of Markdown users and reference manager users:
+only if a manuscript-level Bib(La)TeX file is used the CiTO annotation can be added as notes to the
+BiBTeX file. This is explained in this guidance document: https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html.
 
@@
 
-With X articles published in the CiTO Collection, the adoption is not as high as one would have hoped.
-Nevertheless, the pilot is triggering interest with both authors and publishing platforms. The Scholia
+With fourteen articles published in the CiTO Collection, the adoption is not as high as one would have hoped.
+Nevertheless, the pilot is triggering interest with both authors and publishing platforms. The
+*Journal of Cheminformatics* has already published a few more articles with CiTO annotation and
+a search on the preprint servers ResearchSquare and ChemrXiv show a few more manuscripts. The support
+by BioHackrXiv is a nice example of adoption beyond BioMedCentral.
+
+The Scholia
 use case shows how the information can be used downstream. Further uptake of this idea of typed citations
@@
 #### Availability of data and materials
-Text for this section.
+CiTO annotation in the *Journal of Cheminformatics* are available from the journal's articles. CiTO
+annotation data in Scholia is available from [Wikidata](https://wikidata.org/). Markdown templates that
+support CiTO are available from https://github.com/jcheminform/markdown-jcheminf and
+https://github.com/biohackrxiv/publication-template.
 
@@
 #### Funding
-Text for this section.
+Part of this work was supported by ELIXIR, the research infrastructure for life-science data.
 

### 2022-11-13 article.Rmd
Tweaking references in links

@@
 The *Journal of Cheminformatics* template is available from the journal's GitHub organization, and authors and editors should feel free to adapt it to the needs of other journals.
-The BioHackrXiv (https://biohackrxiv.org/) preprint server also support CiTO annotations [@cito:citesAsDataSource:Prins2021] and this template can be found at https://github.com/biohackrxiv/publication-template.
+The BioHackrXiv ([https://biohackrxiv.org/](https://biohackrxiv.org/))
+preprint server also support CiTO annotations [@citesAsDataSource:Prins2021] and this template can be found at
+[https://github.com/biohackrxiv/publication-template](https://github.com/biohackrxiv/publication-template).
 
@@
 idea which allows giving additional notes to a reference in the bibliography. The CiTO Pilot uses
-the *safe CURIE* standard proposed by the W3C in 2010 [@usesMethodIn:BirBeck2010]. The Markdown templates
+the *safe CURIE* standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010]. The Markdown templates
 implement this idea. However, typesetting in the *Journal of Cheminformatics* has shown to be inconsisten,
@@
 only if a manuscript-level Bib(La)TeX file is used the CiTO annotation can be added as notes to the
-BiBTeX file. This is explained in this guidance document: https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html.
+BiBTeX file. This is explained in this guidance document:
+[https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html](https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html).
 
@@
 
-#### Availability of data and materials
+### Availability of data and materials
+
 CiTO annotation in the *Journal of Cheminformatics* are available from the journal's articles. CiTO
 annotation data in Scholia is available from [Wikidata](https://wikidata.org/). Markdown templates that
-support CiTO are available from https://github.com/jcheminform/markdown-jcheminf and
-https://github.com/biohackrxiv/publication-template.
+support CiTO are available from
+[https://github.com/jcheminform/markdown-jcheminf](https://github.com/jcheminform/markdown-jcheminf) and
+[https://github.com/biohackrxiv/publication-template](https://github.com/biohackrxiv/publication-template).
 
-#### Competing interests
-Text for this section.
+### Competing interests
 
-#### Funding
+No competing interests.
+
+### Funding
+
 Part of this work was supported by ELIXIR, the research infrastructure for life-science data.
 
-#### Authors' contributions
-Text for this section.
+### Authors' contributions
 
-#### Acknowledgements
+Not applicable.
+
+### Acknowledgments
 

### 2022-11-13 article.Rmd
Merged two references

@@
 The Citation Typing Ontology was selected to express the intention [@usesMethodIn:Shotton_2010],
-the intention is expressed a compact identifier wrapped in square brackets [@usesMethodIn:Wimalaratne_2018], and 
-the *bibnotes* concept of the Springer Nature publishing platform was used as carrier.
+the intention is expressed a compact identifier wrapped in square brackets, also called a
+safe CURIE, standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010,@usesMethodIn:Wimalaratne_2018]
+The *bibnotes* concept of the Springer Nature publishing platform was used as carrier.
 Authors are guided by a landing page consisting of a BioMedCentral Collection at
@@
 idea which allows giving additional notes to a reference in the bibliography. The CiTO Pilot uses
-the *safe CURIE* standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010]. The Markdown templates
-implement this idea. However, typesetting in the *Journal of Cheminformatics* has shown to be inconsisten,
-even within single articles, suggesting additional training or guidance could be needed.
+the *safe CURIE* standard, but the typesetting in the *Journal of Cheminformatics* has shown to be inconsistent,
+even within single articles, suggesting additional training or guidance may be needed.
 

### 2022-11-13 article.Rmd
Mention 'cito' is a standardized prefix

@@
 the intention is expressed a compact identifier wrapped in square brackets, also called a
-safe CURIE, standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010,@usesMethodIn:Wimalaratne_2018]
+safe CURIE, standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010,@usesMethodIn:Wimalaratne_2018].
+The *cito* prefix is registered in BioRegistry [@usesDataFrom:Hoyt2022].
 The *bibnotes* concept of the Springer Nature publishing platform was used as carrier.

### 2022-11-13 article.Rmd
Added some more 'next'

@@
 depends on the combined willingness of journal editors, authors, publishers and indexing services alike.
-The rise of services like scite.ai shows that the research community needs this kind of information.
+The rise of services like scite.ai shows that the research community is ready this kind of information.
+Logical steps include support of distributing citation typing annotation via platforms like CrossRef
+or EuropePMC.
+
+Because all innovation requires a critical amount of adoption to be accepted, additional sources of
+CiTO annotation will be welcome. For example, providing CiTO annotation via nanopublications would
+allow collections of annotations to be shared. Then, when archived on Zenodo and therefore citatble,
+these annotations can be included in analysis as a trusted or at least citable source.
 

### 2022-11-13 article.Rmd
Description of a reuse networks, and some layout tweaks

@@
 Citations are an essential aspect of research communication and have become the basis of many
-metrics in the academic publishing world. Some see citation counts as a mark of scientific impact or
-even quality, but in reality the reasons for citing other work are manifold.
-Two years ago the *Journal of Cheminformatics* proposed a pilot for the adoption of annotating
-citations with their annotations. Basically, when you cite a source like a journal
-article or datasets, you also explain why specifically you cite that article. Particularly,
-the agreement and disagreement and reuse of methods and data is interesting.
-This article explores what happened after the launch of the pilot, gives examples of
-how people used the idea, and what needs to happen next.
+metrics in the academic world. Some see citation counts as a mark of scientific impact or
+even quality, but in reality the reasons for citing other work are manifold which makes
+the interpretation more complicated than a single citation count can reflect.
+Two years ago, the *Journal of Cheminformatics* proposed the CiTO Pilot for the adoption of
+a practice of annotating citations with their citation intentions.
+Basically, when you cite a journal article or dataset (or any other soruce), you also explain
+why specifically you cite that source. Particularly,
+the agreement and disagreement and reuse of methods and data are of interest.
+This article explores what happened after the launch of the pilot. We summarize how
+authors in the *Journal of Cheminformatics* used the pilot, shows citation annotations
+are distributed with Wikidata, visualized with Scholia, discusses adoption outside BioMedCentral,
+and ponders about what needs to happen next.
 
@@
 
-## Wikidata and Scholia
+# Wikidata and Scholia
 
@@
 
-![Screenshot of the Scholia Citation Typing Ontology page showing the
-daily statistics.](images/scholiaStats.png)
-
 It must be noted that the *Journal of Cheminformatics* is only one possible source,
@@
 
-## Adoption by *Journal of Cheminformatics* authors
+# Adoption by *Journal of Cheminformatics* authors
 
@@
 far exceeding the original anticipation.
+
+![Screenshot of the Scholia Citation Typing Ontology page showing the
+daily statistics.](images/scholiaStats.png)
 
@@
 
-## Technological innovation
+# Technological innovation
 
@@
 
-## Discussion
+# Annotated Citation Networks
+
+We already use citation networks in finding relevant literature, for example based on co-citation patterns.
+Such analyses become stronger when we know more why articles cited. Similarly, an article that cites an
+article because it uses a method in second article and that method extends a method in a third articles,
+then the first article indirectly uses the method in that third article, even if that third article is
+not directly cited. This is reflecting in citation habits: authors always decide whether to cite all
+articles about a method, only the most recent, or only the oldest (or something else). After all, journals
+frequently have rules about the maximum length of reference sections. Moreover, some methods are
+so well established, we are not expected to cite that work at all.
+
+![Citation network consisting only of citations annotated with **[cito:usesMethodIn]**.
+Articles with explicit CiTO annotation are shown in red. The Wikidata SPARQL query
+used to create this diagram is available from
+[https://gist.github.com/egonw/1c8bc99373a24075838ee19976c74856](https://gist.github.com/egonw/1c8bc99373a24075838ee19976c74856).](images/usesMethodInNetwork.png)
+
+The general availability of open citation allows us to recover such more complex reuse scenarios
+using the citation networks. Moreover, when the citations are annotated, we can zoom in on reuse
+networks. Figure 3 shows a method reuse network for articles with explicit annotation in red. The
+network shows a few article that use methods in two cited articles, e.g. *Kohulan Rajan, 2021*
+and *Henning Otto Brinkhaus, 2022* [@Rajan2021,@Brinkhaus2022]. Of course, if we do not limit
+the network to a subset of articles (here, *Journal of Cheminformatics* articles with CiTO
+annotation), the network becomes more interesting, but also much more complex. Network analysis
+approaches can then be used instead of network visualization.
+
+When we combined the reuse of methods and data, we can for journals summarize which articles
+are most reused. Analyses like this becomes a simple query that can be routinely performed for
+any journal. Figure 4 shows a tabular summary of articles of which methods or data is reused.
+Like citation counts, this data depends totally on explicit citation data. However, these
+citation counts are based on actual reuse and not also on the number of citations as authority.
+
+![Screenshot of the "Most reused articles" section of the Scholia page for the
+*Journal of Cheminformatics*. Based on CiTO annotation available in Wikdiata, the ten
+most reused articles are shown.](images/mostUsedInJCheminf.png)
+
+# Discussion
 
@@
 
-## What is next?
+# What is next?
 

### 2022-11-13 article.Rmd
Fixed typos

@@
 ---
-title: "Two years of CiTO annotations"
-subtitle: "Type: Correspondance"
+title: "Two years of explicit CiTO annotations"
+subtitle: "Type: Correspondence"
 author:
@@
 
-* Correspondance:
+* Correspondence:
 egon.willighagen@maastrichtuniversity.nl
@@
 a practice of annotating citations with their citation intentions.
-Basically, when you cite a journal article or dataset (or any other soruce), you also explain
+Basically, when you cite a journal article or dataset (or any other source), you also explain
 why specifically you cite that source. Particularly,
@@
 
-Communiting new research findings is primarily done by written texts in the form of scholarly
+Communicating new research findings is primarily done by written texts in the form of scholarly
 articles, books, and book chapters. To not having to repeat past research by themselves or others,
@@
 authors cite literature vary, which complicates how we use citations [@agreesWith:citesAsAuthority:Nicholson_2021]. 
-Typing citations is therefore of interest: it allows us to navigate ltierature more easily: it points
+Typing citations is therefore of interest: it allows us to navigate literature more easily: it points
 us to essential research methods, data, and can warn us of research that cannot be reproduced,
@@
 and author guidelines explaining to authors
-how they can add the annotations with their favoriate editor at
+how they can add the annotations with their favorite editor at
 [https://jcheminform.github.io/jcheminform-author-guidelines/cito](https://jcheminform.github.io/jcheminform-author-guidelines/cito).
@@
 and PubMed. Wikidata allows adding qualifier to statements which allowed us to define
-a data model for citations annoted with CiTO intention; the Wikidata property P3712
+a data model for citations annotated with CiTO intention; the Wikidata property P3712
 has been used for this, labeled *objective of project or action* (see Figure 1).
@@
 This data in Wikidata can then be accessed via Scholia and Scholia tells use some
-overall statisticss of the number of annotations, which we reported on about a year
+overall statistics of the number of annotations, which we reported on about a year
 ago too [@citesAsDataSource:Willighagen_2021]. Since last year and written down
@@
 guidance focused on **[cito:citesAsDataSource]**, **[cito:usesDataFrom]** (the first is used
-to cite an article with data, the second when you reused data and citate the article where it
+to cite an article with data, the second when you reused data and cite the article where it
 comes from), **[cito:usesMethodIn]**, **[cito:citesAsAuthority]**, **[cito:discusses]**,
@@
 ![Screenshot of the "Most reused articles" section of the Scholia page for the
-*Journal of Cheminformatics*. Based on CiTO annotation available in Wikdiata, the ten
+*Journal of Cheminformatics*. Based on CiTO annotation available in Wikidata, the ten
 most reused articles are shown.](images/mostUsedInJCheminf.png)
@@
 *Journal of Cheminformatics* has already published a few more articles with CiTO annotation and
-a search on the preprint servers ResearchSquare and ChemrXiv show a few more manuscripts. The support
+a search on the preprint servers ResearchSquare and ChemRxiv show a few more manuscripts. The support
 by BioHackrXiv is a nice example of adoption beyond BioMedCentral.
@@
 CiTO annotation will be welcome. For example, providing CiTO annotation via nanopublications would
-allow collections of annotations to be shared. Then, when archived on Zenodo and therefore citatble,
+allow collections of annotations to be shared. Then, when archived on Zenodo and therefore citable,
 these annotations can be included in analysis as a trusted or at least citable source.
@@
 Matthew Smyllie in particular and the editors of the *Journal of Cheminformatics*,
-Rajarshi Guha, Nina Jealizakova, and Barbara Zdrazil. Carlin MacKenzie is thanked for
+Rajarshi Guha, Nina Jeliazkova, and Barbara Zdrazil. Carlin MacKenzie is thanked for
 integration the sections on CiTO use into the main Scholia journal aspects [@citesAsEvidence:0af60cf7732bc8664d828cbe51f233aa63201df9].

### 2022-11-13 article.Rmd
The Wikidata property is renamed

@@
 a data model for citations annotated with CiTO intention; the Wikidata property P3712
-has been used for this, labeled *objective of project or action* (see Figure 1).
+has been used for this, labeled *objective of project or action* (see Figure 1;
+this property was relabeled in November 2022 as *has goal*).
 

### 2022-11-13 article.Rmd
BMC is th proper name

@@
 authors in the *Journal of Cheminformatics* used the pilot, shows citation annotations
-are distributed with Wikidata, visualized with Scholia, discusses adoption outside BioMedCentral,
+are distributed with Wikidata, visualized with Scholia, discusses adoption outside BMC,
 and ponders about what needs to happen next.
@@
 The *bibnotes* concept of the Springer Nature publishing platform was used as carrier.
-Authors are guided by a landing page consisting of a BioMedCentral Collection at
+Authors are guided by a landing page consisting of a BMC Collection at
 [https://www.biomedcentral.com/collections/cito](https://www.biomedcentral.com/collections/cito)
@@
 a search on the preprint servers ResearchSquare and ChemRxiv show a few more manuscripts. The support
-by BioHackrXiv is a nice example of adoption beyond BioMedCentral.
+by BioHackrXiv is a nice example of adoption beyond BMC.
 

### 2022-11-13 article.Rmd
Renamed the figures according to the submission system

@@
 *objective of project or action* qualifier is used to annotate the citation with their
-CiTO intentions.](images/datamodel.png)
+CiTO intentions.](images/figure1.png)
 
@@
 ![Screenshot of the Scholia Citation Typing Ontology page showing the
-daily statistics.](images/scholiaStats.png)
+daily statistics.](images/figure2.png)
 
@@
 used to create this diagram is available from
-[https://gist.github.com/egonw/1c8bc99373a24075838ee19976c74856](https://gist.github.com/egonw/1c8bc99373a24075838ee19976c74856).](images/usesMethodInNetwork.png)
+[https://gist.github.com/egonw/1c8bc99373a24075838ee19976c74856](https://gist.github.com/egonw/1c8bc99373a24075838ee19976c74856).](images/figure3.png)
 
@@
 *Journal of Cheminformatics*. Based on CiTO annotation available in Wikidata, the ten
-most reused articles are shown.](images/mostUsedInJCheminf.png)
+most reused articles are shown.](images/figure4.png)
 

### 2022-11-24 article.Rmd
s/BioRegistry/Bioregistry/
Added the two requested additional declarations
Removed a bit of reflection on use in the journal

@@
 safe CURIE, standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010,@usesMethodIn:Wimalaratne_2018].
-The *cito* prefix is registered in BioRegistry [@usesDataFrom:Hoyt2022].
+The *cito* prefix is registered in Bioregistry [@usesDataFrom:Hoyt2022].
 The *bibnotes* concept of the Springer Nature publishing platform was used as carrier.
@@
 idea which allows giving additional notes to a reference in the bibliography. The CiTO Pilot uses
-the *safe CURIE* standard, but the typesetting in the *Journal of Cheminformatics* has shown to be inconsistent,
-even within single articles, suggesting additional training or guidance may be needed.
+the *safe CURIE* standard, compatible with the typesetting in the *Journal of Cheminformatics*.
+This makes is easier to text mine the annotations by downstream tools.
 

### 2023-01-08 article.Rmd
Added metadata
Revised the document, clarifying details

@@
 csl: journal-of-cheminformatics.csl
-...
+---
 
@@
 egon.willighagen@maastrichtuniversity.nl
+
+Copyright (c) Egon Willighagen; CC-BY 4.0 International
 
@@
 
-Communicating new research findings is primarily done by written texts in the form of scholarly
+Communicating new research findings is still primarily done by written texts in the form of scholarly
 articles, books, and book chapters. To not having to repeat past research by themselves or others,
-authors cite research relevant [@agreesWith:citesAsAuthority:Shotton_2010]. However, the reasons why
+authors cite relevant research [@agreesWith:citesAsAuthority:Shotton_2010]. However, the reasons why
 authors cite literature vary, which complicates how we use citations [@agreesWith:citesAsAuthority:Nicholson_2021]. 
@@
 us to essential research methods, data, and can warn us of research that cannot be reproduced,
-or others disagree with [@Nicholson_2021].
+or others disagree with [@Nicholson_2021]. Indeed, it helps us understand the history of science
+[@obtainsBackgroundFrom:Park2023].
 
@@
 
-The pilot consists of a couple of components and the editorial explains some of them [@Willighagen_2020].
+The pilot consisted of a couple of components and the editorial explains some of them [@Willighagen_2020].
 The Citation Typing Ontology was selected to express the intention [@usesMethodIn:Shotton_2010],
 the intention is expressed a compact identifier wrapped in square brackets, also called a
-safe CURIE, standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010,@usesMethodIn:Wimalaratne_2018].
+safe CURIE, standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010;@usesMethodIn:Wimalaratne_2018].
 The *cito* prefix is registered in Bioregistry [@usesDataFrom:Hoyt2022].
@@
 The journal itself would adopt the following CiTO annotations: **[cito:retracts]**,
-**[cito:repliesTo]**, and **[updates]**. Fortunately, it has not been used yet, but
+**[cito:repliesTo]**, and **[cito:updates]**. Fortunately, it has not been used yet, but
 the first would be used if an article with be retracted from the journal. The second
@@
 
-This data in Wikidata can then be accessed via Scholia and Scholia tells use some
+This data in Wikidata can then be accessed in multiple ways, including REST APIs and
+an SPARQL interface. The latter is used by Scholia to tell us some
 overall statistics of the number of annotations, which we reported on about a year
-ago too [@citesAsDataSource:Willighagen_2021]. Since last year and written down
-on August 25, the number of annotations and the number of annotated citations
+ago too [@citesAsDataSource:Willighagen_2021]. Since last year and recorded
+on August 25 2022, the number of annotations and the number of annotated citations
 have almost doubled (from 377 to 603 and from 304 to 494, respectively). The first
@@
 
-It must be noted that the *Journal of Cheminformatics* is only one possible source,
-but is still the only journal that uses CiTO annotation explicitly in the articles itself.
+It must be noted that the *Journal of Cheminformatics* is only one possible source
+of CiTO annotations. As far as the author knows, it is still the only journal that uses
+CiTO annotation explicitly in the articles itself. 
 And with 335 annotated citations in 32 articles it also is the major source of
@@
 
-In the first two years after the start of the Pilot, including the seminal editorial [@extends:Willighagen_2020],
-the *Journal of Cheminformatics* has published fifteen articles with explicit CiTO annotation:
+In the two years of the Pilot, including the seminal editorial [@extends:Willighagen_2020],
+the *Journal of Cheminformatics* published fifteen articles with explicit CiTO annotation:
 three Editorials, four Research articles, two Database, and one of each of Data Note, Software,
@@
 
-To make life easier for authors, and following a Twitter discussion in Spring 2021, a Markdown template was developed with native CiTO support: https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html.
-Here, the author merely indicates the CiTO type when they cite the article.
-This is using a method introduced by Krewinkel et al.
-[@usesMethodIn:Krewinkel2017].
-The manuscript can then be converted to a Microsoft Word file with Pandoc (https://pandoc.org/) for submission to the journal.
+To make life easier for authors, and following a Twitter discussion in Spring 2021, a Markdown template was developed with native CiTO support:
+[https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html](https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/markdown.html).
+Here, the author indicates the CiTO type when they cite the article. This is using a method introduced by Krewinkel et al. [@usesMethodIn:Krewinkel2017].
+The manuscript can then be converted to a Microsoft Word file with Pandoc ([https://pandoc.org/](https://pandoc.org/)) for submission to the journal.
+For publishers it will be interesting to note that the Pandoc can be directly converted in
+the *journal article tag suite* (JATS) format [@citesForInformation:Beck2011;@citesAsPotentialSolution:krewinkel2022jats].
 
-The *Journal of Cheminformatics* template is available from the journal's GitHub organization, and authors and editors should feel free to adapt it to the needs of other journals.
-The BioHackrXiv ([https://biohackrxiv.org/](https://biohackrxiv.org/))
-preprint server also support CiTO annotations [@citesAsDataSource:Prins2021] and this template can be found at
+The *Journal of Cheminformatics* template is available from the journal's GitHub organization, and authors and editors
+should feel free to adapt it to the needs of other journals. The BioHackrXiv ([https://biohackrxiv.org/](https://biohackrxiv.org/))
+preprint server also support CiTO annotations [@citesAsDataSource:citesAsPotentialSolution:Prins2021] and this template can be found at
 [https://github.com/biohackrxiv/publication-template](https://github.com/biohackrxiv/publication-template).
@@
 network shows a few article that use methods in two cited articles, e.g. *Kohulan Rajan, 2021*
-and *Henning Otto Brinkhaus, 2022* [@Rajan2021,@Brinkhaus2022]. Of course, if we do not limit
+and *Henning Otto Brinkhaus, 2022* [@Rajan2021;@Brinkhaus2022]. Of course, if we do not limit
 the network to a subset of articles (here, *Journal of Cheminformatics* articles with CiTO
@@
 Because many publishing platforms currently do not support display of citation level intention annotation,
-the default model only provides the annotation in the bibliography. The choice was made to use the `bibnotes`
-idea which allows giving additional notes to a reference in the bibliography. The CiTO Pilot uses
+the simplest model only provides the annotation in the bibliography. The Pilot made this choice to
+be able to use the `bibnotes` approach which allows giving additional notes to a reference in the bibliography. The CiTO Pilot uses
 the *safe CURIE* standard, compatible with the typesetting in the *Journal of Cheminformatics*.
-This makes is easier to text mine the annotations by downstream tools.
+This makes is easier to text mine the annotations by downstream tools, from both the HTML and PDF as
+well as the JATS versions of the article.
 
-This links to a limitation to the current use of CiTO annotation: the citation-level annotation may be
+However, this links also to the limitation to the current use of CiTO annotation: the citation-level annotation may be
 supported by some authoring systems (Markdown), even then the depiction may not be supported. If we convert
@@
 only if a manuscript-level Bib(La)TeX file is used the CiTO annotation can be added as notes to the
-BiBTeX file. This is explained in this guidance document:
+BiBTeX file. This way, the CiTO annotation is specific for this manuscript and each manuscript can use
+different annotations. This approach is explained in this guidance document:
 [https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html](https://jcheminform.github.io/jcheminform-author-guidelines/cito-guidelines/latex.html).
+
+From a use case perspective, it is easy to see how this kind of annotation can be used. For example,
+we here showed examples of reuse of work. via **[[cito:usesMethodIn]** and **[cito:usesDataFrom]**.
+Second, scite.ai is a clear use case of **[cito:agreesWith]** and **[cito:disagreesWith]** annotation,
+though it makes an good case how such citation intentions can be extracted with text mining instead.
+However, other use cases still need development and adoption, which brings us to the question:
+what is next?
 
@@
 
-With fourteen articles published in the CiTO Collection, the adoption is not as high as one would have hoped.
-Nevertheless, the pilot is triggering interest with both authors and publishing platforms. The
-*Journal of Cheminformatics* has already published a few more articles with CiTO annotation and
-a search on the preprint servers ResearchSquare and ChemRxiv show a few more manuscripts. The support
-by BioHackrXiv is a nice example of adoption beyond BMC.
+With fifteen articles published in the CiTO Collection, the pilot triggered interest from authors.
+The *Journal of Cheminformatics* has already published a few more articles with CiTO annotation
+after August 2022 and a search on the preprint servers ResearchSquare and ChemRxiv [@citesAsAuthority:Mudrak2022] show a
+few more manuscripts. The support by BioHackrXiv [@citesAsAuthority:Prins2022] is a nice example of
+adoption beyond BMC. Further citation intention annotations will come from literature studies where
+citations networks are characterized. For example, Duca *et al.* used CiTO annotations to describe
+the citation network to retracted COVID-19 articles [@citesAsRecommendedReading:Duca2022].
 
-The Scholia
-use case shows how the information can be used downstream. Further uptake of this idea of typed citations
+Further uptake of this idea of typed citations
 depends on the combined willingness of journal editors, authors, publishers and indexing services alike.
 The rise of services like scite.ai shows that the research community is ready this kind of information.
-Logical steps include support of distributing citation typing annotation via platforms like CrossRef
-or EuropePMC.
+Logical steps include support of distributing citation typing annotation via platforms like PlumX,
+Altmetric.com, CrossRef, or EuropePMC. Support of CiTO annotation in the JATS format seems an
+important step forward.
 
 Because all innovation requires a critical amount of adoption to be accepted, additional sources of
-CiTO annotation will be welcome. For example, providing CiTO annotation via nanopublications would
-allow collections of annotations to be shared. Then, when archived on Zenodo and therefore citable,
+CiTO annotation will be welcome. For example, providing CiTO annotations via a standardized format
+like a spreadsheet-like format or nanopublications would allow collections of annotations to be shared,
+such as that by Duca *et al.*. Then, when archived on Zenodo and therefore citable,
 these annotations can be included in analysis as a trusted or at least citable source.
@@
 [https://github.com/biohackrxiv/publication-template](https://github.com/biohackrxiv/publication-template).
+An archive of all CiTO annotations in Wikidata is available from [https://doi.org/10.5281/zenodo.7513573](https://doi.org/10.5281/zenodo.7513573).
 

### 2023-01-08 article.Rmd
Added some non-breaking spaces

@@
 articles, books, and book chapters. To not having to repeat past research by themselves or others,
-authors cite relevant research [@agreesWith:citesAsAuthority:Shotton_2010]. However, the reasons why
-authors cite literature vary, which complicates how we use citations [@agreesWith:citesAsAuthority:Nicholson_2021]. 
+authors cite relevant research&nbsp;[@agreesWith:citesAsAuthority:Shotton_2010]. However, the reasons why
+authors cite literature vary, which complicates how we use citations&nbsp;[@agreesWith:citesAsAuthority:Nicholson_2021]. 
 Typing citations is therefore of interest: it allows us to navigate literature more easily: it points
 us to essential research methods, data, and can warn us of research that cannot be reproduced,
-or others disagree with [@Nicholson_2021]. Indeed, it helps us understand the history of science
-[@obtainsBackgroundFrom:Park2023].
+or others disagree with&nbsp;[@Nicholson_2021]. Indeed, it helps us understand the history of
+science&nbsp;[@obtainsBackgroundFrom:Park2023].
 
 With the use of citations increasingly being picked up to help researchers with tools like
-scite.ai [@citesAsAuthority:Nicholson_2021] en Connected Papers [@citesAsAuthority:ConnectedPapers], having typed
+scite.ai&nbsp;[@citesAsAuthority:Nicholson_2021] en Connected Papers&nbsp;[@citesAsAuthority:ConnectedPapers], having typed
 citations will help us explore literature. Therefore, the *Journal of Cheminformatics*
-started a pilot for the adoption of annotating citations with their annotations [@discusses:Willighagen_2020].
+started a pilot for the adoption of annotating citations with their annotations&nbsp;[@discusses:Willighagen_2020].
 
@@
 
-The pilot consisted of a couple of components and the editorial explains some of them [@Willighagen_2020].
-The Citation Typing Ontology was selected to express the intention [@usesMethodIn:Shotton_2010],
+The pilot consisted of a couple of components and the editorial explains some of them&nbsp;[@Willighagen_2020].
+The Citation Typing Ontology was selected to express the intention&nbsp;[@usesMethodIn:Shotton_2010],
 the intention is expressed a compact identifier wrapped in square brackets, also called a
-safe CURIE, standard proposed by the W3C in 2010 [@usesMethodIn:McCarron2010;@usesMethodIn:Wimalaratne_2018].
-The *cito* prefix is registered in Bioregistry [@usesDataFrom:Hoyt2022].
+safe CURIE, standard proposed by the W3C in 2010&nbsp;[@usesMethodIn:McCarron2010;@usesMethodIn:Wimalaratne_2018].
+The *cito* prefix is registered in Bioregistry&nbsp;[@usesDataFrom:Hoyt2022].
 The *bibnotes* concept of the Springer Nature publishing platform was used as carrier.
@@
 Because the CiTO ontology has many terms for many different citation intentions,
-we made a selection of CiTO terms authors could use [@CiTOCollection]:
+we made a selection of CiTO terms authors could use&nbsp;[@CiTOCollection]:
 **[cito:citesAsDatasource]** to indicate
@@
 citation intention data. Scholia is a graphical interface around the data stored in
-Wikidata [@usesMethodIn:Nielsen_2017] and includes citations from OpenCitations [@Peroni_2020]
+Wikidata&nbsp;[@usesMethodIn:Nielsen_2017] and includes citations from OpenCitations&nbsp;[@Peroni_2020]
 and PubMed. Wikidata allows adding qualifier to statements which allowed us to define
 a data model for citations annotated with CiTO intention; the Wikidata property P3712
-has been used for this, labeled *objective of project or action* (see Figure 1;
+has been used for this, labeled *objective of project or action* (see Figure&nbsp;1;
 this property was relabeled in November 2022 as *has goal*).
@@
 overall statistics of the number of annotations, which we reported on about a year
-ago too [@citesAsDataSource:Willighagen_2021]. Since last year and recorded
+ago too&nbsp;[@citesAsDataSource:Willighagen_2021]. Since last year and recorded
 on August 25 2022, the number of annotations and the number of annotated citations
@@
 scholarly journals, and they are found in 98 articles in 48 different journals
-(see Figure 2) [@citesAsDataSource:scholiaCiTOStatistics].
+(see Figure&nbsp;2)&nbsp;[@citesAsDataSource:scholiaCiTOStatistics].
 
@@
 Letter to the Editor, Letter Response, Educational, and Methodology. Ten were published in
-the first year (Table 1) and five in the second year (Table 2). Each article annotated
+the first year (Table&nbsp;1) and five in the second year (Table&nbsp;2). Each article annotated
 one or more citations with CiTO intentions, and several articles annotated every citation,
@@
 using the citation networks. Moreover, when the citations are annotated, we can zoom in on reuse
-networks. Figure 3 shows a method reuse network for articles with explicit annotation in red. The
+networks. Figure&nbsp;3 shows a method reuse network for articles with explicit annotation in red. The
 network shows a few article that use methods in two cited articles, e.g. *Kohulan Rajan, 2021*
@@
 are most reused. Analyses like this becomes a simple query that can be routinely performed for
-any journal. Figure 4 shows a tabular summary of articles of which methods or data is reused.
+any journal. Figure&nbsp;4 shows a tabular summary of articles of which methods or data is reused.
 Like citation counts, this data depends totally on explicit citation data. However, these
@@
 From a use case perspective, it is easy to see how this kind of annotation can be used. For example,
-we here showed examples of reuse of work. via **[[cito:usesMethodIn]** and **[cito:usesDataFrom]**.
+we here showed examples of reuse of work, via **[cito:usesMethodIn]** and **[cito:usesDataFrom]**.
 Second, scite.ai is a clear use case of **[cito:agreesWith]** and **[cito:disagreesWith]** annotation,
@@
 Rajarshi Guha, Nina Jeliazkova, and Barbara Zdrazil. Carlin MacKenzie is thanked for
-integration the sections on CiTO use into the main Scholia journal aspects [@citesAsEvidence:0af60cf7732bc8664d828cbe51f233aa63201df9].
+integration the sections on CiTO use into the main Scholia journal aspects&nbsp;[@citesAsEvidence:0af60cf7732bc8664d828cbe51f233aa63201df9].
 Huge thanks goes to Albert Krewinkel for developing the Markdown/Pandoc integration.

### 2023-01-09 article.Rmd
Applied suggestions by Rajarshi

@@
 Citations are an essential aspect of research communication and have become the basis of many
-metrics in the academic world. Some see citation counts as a mark of scientific impact or
+evaluation metrics in the academic world. Some see citation counts as a mark of scientific impact or
 even quality, but in reality the reasons for citing other work are manifold which makes
@@
 are distributed with Wikidata, visualized with Scholia, discusses adoption outside BMC,
-and ponders about what needs to happen next.
+and finally present some thoughts on what needs to happen next.
 
@@
 citations will help us explore literature. Therefore, the *Journal of Cheminformatics*
-started a pilot for the adoption of annotating citations with their annotations&nbsp;[@discusses:Willighagen_2020].
+started a pilot to explore capturing the intent of citations using annotations&nbsp;[@discusses:Willighagen_2020].
 
@@
 
-# Adoption by *Journal of Cheminformatics* authors
+# Adoption by *Journal of Cheminformatics* Authors
 
@@
 
-# Technological innovation
+# Technological Innovation
 
@@
 For publishers it will be interesting to note that the Pandoc can be directly converted in
-the *journal article tag suite* (JATS) format [@citesForInformation:Beck2011;@citesAsPotentialSolution:krewinkel2022jats].
+the *Journal Article Tag Suite* (JATS) format [@citesForInformation:Beck2011;@citesAsPotentialSolution:krewinkel2022jats].
 
@@
 
-# What is next?
+# What is Next?
 

### 2023-01-13 article.Rmd
Applied suggestions from Barbara

@@
 **[cito:citesAsDatasource]** to indicate
-to a source the provides data to back up a claim, **[cito:usesDataFrom]** to indicate that the authors
-reused data, **[cito:usesMethodIn]** when a method or protocol explain in that source is used,
+a source that provides data to back up a claim, **[cito:usesDataFrom]** to indicate that the authors
+reused data, **[cito:usesMethodIn]** when a method or protocol explained in that source is used,
 and a few more general intentions like **[cito:discusses]**, **[cito:extends]**, **[cito:agreesWith]**,
@@
 **[cito:repliesTo]**, and **[cito:updates]**. Fortunately, it has not been used yet, but
-the first would be used if an article with be retracted from the journal. The second
-would be used when an Letter to the Editor replies to an earlier published article,
-and **[cito:updates]** when an Correction was published.
+the first would be used if an article was retracted from the journal. The second
+would be used when a Letter to the Editor replies to an earlier published article,
+and **[cito:updates]** when a Correction was published.
 
@@
 Wikidata&nbsp;[@usesMethodIn:Nielsen_2017] and includes citations from OpenCitations&nbsp;[@Peroni_2020]
-and PubMed. Wikidata allows adding qualifier to statements which allowed us to define
+and PubMed. Wikidata allows adding qualifiers to statements which allowed us to define
 a data model for citations annotated with CiTO intention; the Wikidata property P3712
@@
 This data in Wikidata can then be accessed in multiple ways, including REST APIs and
-an SPARQL interface. The latter is used by Scholia to tell us some
+a SPARQL interface. The latter is used by Scholia to tell us some
 overall statistics of the number of annotations, which we reported on about a year
@@
 then the first article indirectly uses the method in that third article, even if that third article is
-not directly cited. This is reflecting in citation habits: authors always decide whether to cite all
+not directly cited. This is reflecting citation habits: authors always decide whether to cite all
 articles about a method, only the most recent, or only the oldest (or something else). After all, journals
@@
 
-When we combined the reuse of methods and data, we can for journals summarize which articles
-are most reused. Analyses like this becomes a simple query that can be routinely performed for
+When we combine the reuse of methods and data, we can for journals summarize which articles
+are most reused. Analyses like this become a simple query that can be routinely performed for
 any journal. Figure&nbsp;4 shows a tabular summary of articles of which methods or data is reused.
@@
 Second, scite.ai is a clear use case of **[cito:agreesWith]** and **[cito:disagreesWith]** annotation,
-though it makes an good case how such citation intentions can be extracted with text mining instead.
+though it makes a good case how such citation intentions can be extracted with text mining instead.
 However, other use cases still need development and adoption, which brings us to the question:
@@
 depends on the combined willingness of journal editors, authors, publishers and indexing services alike.
-The rise of services like scite.ai shows that the research community is ready this kind of information.
-Logical steps include support of distributing citation typing annotation via platforms like PlumX,
-Altmetric.com, CrossRef, or EuropePMC. Support of CiTO annotation in the JATS format seems an
-important step forward.
+The rise of services like scite.ai shows that the research community is ready for this kind of information.
+Logical steps forward include support of distributing citation typing annotation via platforms like PlumX,
+Altmetric.com, CrossRef, or EuropePMC, and support of CiTO annotation in the JATS format.
 
 Because all innovation requires a critical amount of adoption to be accepted, additional sources of
-CiTO annotation will be welcome. For example, providing CiTO annotations via a standardized format
+CiTO annotation will be welcomed. For example, providing CiTO annotations via a standardized format
 like a spreadsheet-like format or nanopublications would allow collections of annotations to be shared,
